Rice yield improvements by plant breeding are offset by decreases over time

### 2017-07-03 output/breed_paper.Rnw
Update sequence of paper

\newcommand{\tsups}{\textsuperscript}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report}

@@

Meeting the challenge of feeding an growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite demonstrated increased yield
potential of new genetics over old in side-by-side comparisons. We
analyzed experimental data from one such system, California (CA) rice
production, in order to better understand what improvements have been
made in the genetic yield potential via plant breeding. Specifically,
the hypothesis that a rice variety's yield erodes over time after the
initial selection was tested as an explanation for how side-by-side
comparisons show increased yield while no substantive increases have
been seen in farmers' fields. We confirmed an apparent increase in
yield in side-by-side comparisons between old and newer varieties when
the difference between the years since release was not
considered. However, we demonstrate this apparent increase in yield
can be explained by the differences in the time since selection
between cultivars. The impact of yield erosion was estimated to be
29.8 kg ha\tsups{-1} year\tsups{-1} after initial
selection. Once this effect was considered, the apparent yield
advantage of newer cultivars over old was uncertain. This underscores
the importance of breeding to maintain yield and the challenges of
increasing the genetic yield potential.

@@

id_est = ranef(RES_full)$id
id_est = cbind(id_est, posterior_interval(RES_full, regex_pars = "id:M"))
idx = order(id_est[,1], decreasing = FALSE)
id_est = id_est[idx,]
@

@@

\begin{figure}[H]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
                  CA_data$Value * 1.12,
                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
                rep(1/11,11))
lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  -- \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
  US Dept. of Agriculture National Agricultural Statistics
  Service. Rapid gains in grain yield were experienced from
  approximately 1950 -- 1990, but annual gains since 1990 have been
  slower. The line represents the 10 year moving average.}
\label{fig:state_yield}
\end{figure}

@@

As part of the annual Statewide Variety Trials, newly developed
entries are tested against named and released cultivars (checks) at
the RES in multiple trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute roughly
90\% of the planted area in CA. For these cultivars planted in
breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations were
collected. This lead to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release (typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability) and the
time elapse (in years) since release was calculated.

Plots at the RES are managed similar to prescribed best management
practice, including land preparation, fertility management, and pest
and disease control. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

% \begin{equation}
%  yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
%    release\ i} + \beta_{release\ year\ i}*x_{release\ year\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
% \end{equation}

% Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
% average yield for the RES across varieties and years (intercept),
% $\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
% release\ i}$, the number of years $i$'s cultivar has been in the
% trials (negative for years prior to release, zero at the year of
% release), $\beta_{release\ year\ i}$ is the effect of $x_{release\
% year\ i}$, the year of release for $i$'s cultivar, $\beta_{year}$ is
% the random effect of year, and $\beta_{cultivar}$ is the random effect
% of cultivar.

Data was processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo},
an efficient means of conducting leave-one-out cross validation
\citep{loo_paper}, with the full model estimated to have slightly
better predictive accuracy compared to the reduced models.

\section{Results}\label{results}

@@

There was no conclusive evidence that the yields of rice cultivars
selected by CCRRF have, on average, increased over the last 30
years. The estimated change in grain yield per year was estimated to
be a negligible \Sexpr{release_yr_coef} kg ha\tsups{-1}
year\tsups{-1}, with both small positive and negative trends within
the credible interval (90\% CI: \Sexpr{rel_CI[1]} --
\Sexpr{rel_CI[2]}). Hence, the evidence here is consistent with little
to no increase in yield potential over the study period. When yield
erosion is not corrected for in the comparison (i.e., omitted from the
model), there is an apparent small increase
per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}), though this
estimate ignores the confounding effect of yield erosion over
time.

There do appear to be some differences between cultivars, with some
cultivars showing higher yield than the overall average. The marginal
estimate for the performance of a cultivar (averaged over the
uncertainty in the other parameters in the model) shows three
cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=', ')}
(Fig. \ref{fig:ids_comp}). A fourth, M-209, might also have higher
than average yield, though the CI overlaps with zero.

@@

res = mod_data$site == "RES"
comp_yrs = 2010:2015
idx = mod_data$year %in% comp_yrs

@@

with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
mtext("a.", adj = 0)

plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
     yaxt = "n",
     ylim = c(8, 12.5),
     xaxt = "n",
     xlim = c(1980,2020))
axis(1, at = seq(1980, 2020, by = 10))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
                         y1 = mean - se))
text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
     labels = gsub("M", "M-", row.names(plot_data2)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
mtext("b.", adj = 0)
polygon(x = c(yrs, rev(yrs)),
        y = c(ints2[1,],rev(ints2[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

@
\caption{The average yield ($\bullet$) and standard error (|) of
medium grain rice cultivars by year of release (a.) without correcting
for yield erosion over time and (b.) accounting for yield erosion over
time. The trend without correcting for yield erosion (a.) was estimated by
omitting the year of each cultivar's release from the
model. Yield erosion was corrected for by including the years since
release (b.) in the model. predictions assuming no yield erosion
were generated from the model which included both the year of release
and the years since release. For display purposes, only data for
experimental plots planted at the California Cooperative Rice Research
Foundation Rice Experiment Station (California, USA) over the years
\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)} are included.}
\label{fig:release}
\end{figure}

\subsection{Yield decreases over time}

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is
  estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
  year\tsups{-1}. Points ($\bullet$) and bars (|) represent the mean and standard
  error of the raw data, respectively. The
  shaded area is the 90\% credible interval.}
\label{fig:RES_year}
\end{figure}

There is estimated to be a small loss of yield over time once a
cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}, 90\% CI:
\Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in the
trials is estimated to have lost \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since release. Comparing two dominate
medium grain rice cultivars (M-202, which was widely planted from the late 1980s until
the early 2000s, and M-206, the most widely planted CRRFS
cultivar at approximately 50\% of planted area at the time of
writing), the two cultivars experienced similar trends over time to
each other (Fig. \ref{fig:compare_yrs}) and to the estimate from all
released medium grains (Table \ref{tbl:coef}). There is an estimated
\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
over M-202 at the time of writing due solely to the erosion of M-202's
yield over time (Figs \ref{fig:release} \&
\ref{fig:compare_vars}). Further supporting this estimate, both cultivars yielded
similarly at their respective year of release
(Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was released
19 years prior to M-206 (Fig. \ref{fig:release}, Table \ref{tbl:ids}).

@@

Overall, we do not find conclusive evidence of increased yield
potential increasing in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons
for yield erosion over time in the older cultivar, the estimated yield
improvement spans from \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg
ha\tsups{-1} year\tsups{-1}.  Even when the
comparison is not corrected for differences in years since release
between cultivars, the estimated improvements over time
(\Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; Fig. \ref{fig:release} dashed line) are
still much less than previous estimates of 37 -- 60 kg
ha\tsups{-1} year\tsups{-1}
\citep{Omar2014}. Given this evidence, we conclude that the breeding
program has likely been maintaining but not substantially improving
yield potential in the face of changing biotic and abiotic
pressures. These biotic and abiotic forces are expected to erode yield
over time \citep{Peng1998, Peng2000, Peng2010}. In fact, the trend
seen here is extremely consistent with previous reports of erosion of
genetic gains of cultivar IR8 at IRRI \citep{Peng1998, Peng2000}. The
evidence here further supports that the erosion of yield over time is
a phenomenon that affects all rice breeding programs. This further
highlights both the need for maintenance breeding and the difficulty
of increasing genetic yield potential as yield approaches the
physiological limit. Without maintenance breeding continuously
selecting cultivars well adapted to current conditions, CA rice
production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1}
grain yield over the study period.

The goal of a breeding program is to not only maintain the
benchmark yield of the previous ``standard'' but to surpass it. Given
the estimated yield decreases over time, to hit an ambitious target of
0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield
per release (with roughly 4 -- 5 years between releases), a new cultivar
needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in
side-by-side comparison on top of the 0.5 Mg ha\tsups{-1}
target improvement. Often realized side-by-side yield
improvements 
are closer to 2 -- 3\% per release, in which case no absolute
improvements are made beyond maintenance. The evidence here is
consistent with the latter situation occurring in CA, further suggesting
that the genetic yield potential has not increased substantially in
the last 30+ years (Fig. \ref{fig:release}).

@@

That is not to say there have not been yield improvements over the
last 30 years. California state average yields have increased slowly,
even as genetic gains have been small
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management \citep{Lundy2012,
Lundy2015, linquist2009}, weed management \citep{Saito2008,
Pittelkow2012, linquist2008}, etc. may all be contributing to
increases in the state average yield. However, these gains are less
than those experienced during the ``Green revolution'', and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers, state
average yields (approximately 10 Mg ha\tsups{-1} at the time
of writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

@@

Improvement in grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. To achieve these
improvements, we show here that (1) continuous maintenance via plant
improvement is required, without which yields would decrease over
time, (2) the challenge of improving genetic yield potential is
substantial, since absolute gains require improvement in yield in
excess of the maintenance threshold, and (3) increasing yield through
improved management (agronomy) cannot exceed the ceiling of genetic
yield potential. Taken together, these conclusions underscore both the
necessity and substantial challenges facing rice breeding. We suggest
here that historical benchmarks be used in to evaluate promising rice
cultivars in addition to side-by-side comparisons to maintain
perspective of the absolute yield gains being made.

@@

CIs = posterior_interval(RES_full)[1:3,] * scl
CIs[1,] = round(CIs[1,] + center, -2)
CIs = round(CIs, 1)
med_eff = fixef(RES_full) * scl
med_eff[1] = round(med_eff[1] + center, -2)
med_eff = round(med_eff, 1)

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
rownames(RES_coef) = c("Intercept (kg ha\\tsups{-1})",
                       "Release year (kg ha\\tsups{-1} year\\tsups{-1})",
                       "Years since release (kg ha\\tsups{-1} year\\tsups{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

print(xtable(RES_coef,
             caption = "Estimated median effects and 90\\% credible intervals.",
             label = "tbl:coef",
             align = "lcc"),
      caption.placement="top",
      booktabs=TRUE,
      sanitize.colnames.function = function(x){x},
      sanitize.rownames.function = function(x){x})
@
\newpage

>>>>

\newcommand{\tsups}{\textsuperscript}
\usepackage{Sweave}
\SweaveOpts{prefix.string=graphics/report, png=TRUE}

@@

Meeting the challenge of feeding an growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite demonstrated increased yield
potential of new genetics over old in side-by-side comparisons. We
analyzed experimental data from one such system, California (CA) rice
production, in order to better understand what improvements have been
made in the genetic yield potential via plant breeding. Specifically,
the hypothesis that a rice variety's yield erodes over time after the
initial selection was tested as an explanation for how side-by-side
comparisons show increased yield while no substantive increases have
been seen in farmers' fields. We confirmed an apparent increase in
yield in side-by-side comparisons between old and newer varieties when
the yield erosion was not
considered. However, we demonstrate this apparent increase in yield
can be explained by the differences in the time since selection
between cultivars. The impact of yield erosion was estimated to be
29.8 kg ha\tsups{-1} year\tsups{-1} after initial
selection. Once this effect was considered, the apparent yield
advantage of newer cultivars over old was uncertain. This underscores
the importance of breeding to maintain yield and the challenges of
increasing the genetic yield potential.

@@

id_est = ranef(RES_full)$id
id_est = cbind(id_est, posterior_interval(RES_full, regex_pars = "id:M"))
idx = order(id_est[,1], decreasing = FALSE)
id_est = id_est[idx,]

@

@@

\begin{figure}[H]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
# mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
#                   CA_data$Value * 1.12,
#                   rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
#                 rep(1/11,11))
# lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
  -- \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
  US Dept. of Agriculture National Agricultural Statistics
  Service. Rapid gains in grain yield were experienced from
  approximately 1950 -- 1990, but annual gains since 1990 have been
  slower.}%The line represents the 10 year moving average.}
% Remove 10 year moving average - Bruce does not like. 
\label{fig:state_yield}
\end{figure}

@@

As part of the annual Statewide Variety Trials, newly developed
entries are tested against named and released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute roughly
90\% of the planted area in CA. These cultivars are typcically planted
in the Statewide Variety Trials for 2 -- 4 years prior to their
official release. For these cultivars planted in
breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations were
collected leading to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release (typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability) and the
time elapsed (in years) since release was calculated.

Plots at the RES are managed similar to University of California
Cooperative Extension prescribed best management
practice, including land preparation, fertility management, and pest
and disease control \citep{ucce}. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

\begin{equation}
\begin{split}
 yield_i = Intercept + (Rate_{yield\ erosion\ i}* {years\ since\
   release\ i}) \\+ (Rate_{yield\ improvement\ i}* {release\ year\ i}) 
 + \beta_{year\ i} + \beta_{cultivar\ i} \\
\end{split}
\end{equation}

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across varieties and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was as above except omitting the $Rate_{yield\ erosion\ i}$ term.

% Add more detail to the methods section - tell how you get these
% numbers. "Evidence for erosion" - what does this mean for yield
% increases? Erosion 1st, 2nd implications of erosion. 
% Drop off -5 and -6

Data was processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval \citep{rstanarm}. Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

\section{Results}\label{results}
% Invert this section, then talk about yield improvement accounting
% for yield erosion.

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated erosion of yield over time
  (averaged over differences in cultivar and year) is estimated to be
  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
  and bars (|) represent the mean and standard error of the raw data,
  respectively. The shaded area is the 90\% credible interval.}
% For each year, point size by n (open circles)
\label{fig:RES_year}
\end{figure}

\subsection{Yield erosion}

There is estimated to be an decrease of yield over time once a
cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}, 90\% CI: \Sexpr{RES_CI[2]} --
\Sexpr{RES_CI[1]}). The oldest cultivar in the trials is estimated to
have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since release. Comparing two dominate medium grain rice
cultivars M-202 (widely planted from the late 1980s until the early
2000s), and M-206 (the most widely planted CCRRFS cultivar at
approximately 50\% of planted area at the time of writing), the two
cultivars experienced similar trends over time to each other
(Fig. \ref{fig:compare_yrs}) and to the estimate from all released
medium grains (Table \ref{tbl:coef}). At the time of writing, an
estimated \Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
over M-202 can be attributed solely to the erosion of M-202's yield
over time (Figs \ref{fig:release} \& \ref{fig:compare_vars}). Further
supporting this estimate, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:release}, Table \ref{tbl:ids}).

@@

The full model which corrected for both the year of release and yield
erosion was estimated to have better out-of-sample predictive accuracy
compared to reduced models containing only one predictor. As estimated
by the best model, there is no conclusive evidence that the yields of
medium grain rice cultivars selected and released by CCRRF's breeding
program have, on average, increased over the last 30 years. The
estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was estimated to be a negligible
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
(Fig. \ref{fig:release}b.). Hence, the evidence here is consistent
with little to no increase in yield potential over the study
period. When yield erosion is not corrected for in the comparison
(i.e., omitted from the model), there is an apparent small increase
per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}a.), though this
estimate ignores the confounding effect of yield erosion over time and
performed worse in out-of-sample predictive accuracy.

After correcting for the other variables, some cultivars are estimated
to have greater yield compared to the overall average. The marginal
estimate for the performance of a cultivar averaged over the
uncertainty in the other parameters in the model (i.e., the random
effect) shows three cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:release} \& \ref{fig:ids_comp}). A fourth, M-209,
might also have higher than average yield, though the CI for M-209
overlaps with zero (Fig. \ref{fig:ids_comp}).

@@

res = mod_data$site == "RES"
comp_yrs = 2011:2016
idx = mod_data$year %in% comp_yrs

@@

with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
mtext("Without adjusting for yield erosion", adj = 0)

plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
     yaxt = "n",
     ylim = c(8, 12.5),
     xaxt = "n",
     xlim = c(1980,2020))
axis(1, at = seq(1980, 2020, by = 10))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
                         y1 = mean - se))
text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
     labels = gsub("M", "M-", row.names(plot_data2)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
mtext("Adjusting for yield erosion", adj = 0)
polygon(x = c(yrs, rev(yrs)),
        y = c(ints2[1,],rev(ints2[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

@
% Change from a./b. to accounting for erosion, More details about how
% yield was corrected - 
\caption{The relationship between yield and the release year of medium
  grain rice cultivars developed by the California Cooperative Rice
  Research Foundation (California, USA), for the last 6 years (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by two methods;
  without correcting for yield erosion over time (left) and 
  correcting for yield erosion over time (right). The trend without correcting
  for yield erosion (left) was estimated by omitting the year of each
  cultivar's release from the model. Yield erosion was corrected for by
  including the years since release (right) in the model.}
\label{fig:release}
\end{figure}

@@

We conclude that yield erosion has decreased the yield of cultivars
released by the CCRRF over the last 30+ years. Biotic and abiotic
forces have been speculated to be responsible for this
yield erosion since the cultivar's genetics remain
``frozen'' at the time of selection but the conditions the cultivar
may experience over time are continuously changing \citep{Peng1998,
Peng2000, Peng2010}. The trend observed here is consistent with
previous reports of the erosion of genetic gains of cultivar IR8 at
IRRI \citep{Peng1998, Peng2000}, suggesting the
erosion of yield over time is a phenomenon that widely affects rice
breeding programs. This highlights both the need for maintenance
breeding and the difficulty of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit \citep{espe16a}. Without maintenance breeding
continuously selecting cultivars well adapted to current conditions,
CA rice production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1} grain yield
over the study period.

Overall, we do not find conclusive evidence of increased genetic yield
potential in medium grain rice cultivars released by CCRRF's breeding
improvement program over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons for
yield erosion over time, the estimated yield improvement spans from
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}.
Even when the comparison is not corrected for yield erosion, the
estimated improvements over time (\Sexpr{only_release_coef} kg
ha\tsups{-1} year\tsups{-1}; Fig. \ref{fig:release}a.) are still much
less than previous estimates of 37 -- 60 kg ha\tsups{-1}
year\tsups{-1} \citep{Omar2014}. We conclude that the breeding program
has likely been maintaining but not substantially improving yield
potential in the face of changing biotic and abiotic pressures.

However, the goal of a breeding program is to not only maintain the
benchmark yield of the previous ``standard'' but to surpass it. Given
the estimated yield decreases over time, to hit an ambitious target of
0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield per release
(with roughly 4 -- 5 years between releases), a new cultivar needs to
yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in side-by-side
comparison on top of the 0.5 Mg ha\tsups{-1} target improvement. Often
realized side-by-side yield improvements are closer to 2 -- 3\% per
release, in which case no absolute improvements are made beyond
maintenance. The evidence here is consistent with the latter situation
occurring in CA, further suggesting that the genetic yield potential
has not increased substantially in the last 30+ years
(Fig. \ref{fig:release}).

@@

California state average yields have increased slowly
even as little to no gains have been seen in the breeding trials
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible for these gains
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management \citep{Lundy2012,
Lundy2015, linquist2009}, weed management \citep{Saito2008,
Pittelkow2012, linquist2008}, etc. may all be contributing to
increases in the state average yield. However, these gains are less
than those experienced during the ``Green revolution'', and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers' fields, state
average yields (approximately 10 Mg ha\tsups{-1} at the time of
writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

There are several cultivars that were estimated to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:comp_ids})) may not be widely adapted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields \citep{espe2017}, and it is also known that the RES
experiences warmer temperatures than many other rice growing areas in
CA \citep{espe2017}. These cultivars may perform well at the RES but not widely in
CA. Therefore, it should be noted that while our investigation has
been concerned about increasing yield potential, the estimates here
might not capture yield stability. Yield stability is notoriously
difficult to quantify (CITE), hence further investigation is needed to
explore how yield stability has changed over the study period.

@@

Improvement in grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. To achieve these
improvements, we discuss here that (1) continuous maintenance via plant
improvement is required, without which yields would decrease over
time, (2) the challenge of improving genetic yield potential is
substantial, since absolute gains require improvement in yield in
excess of the maintenance threshold, and (3) increasing yield through
improved management (agronomy) cannot exceed the ceiling of genetic
yield potential. Taken together, these conclusions underscore both the
necessity and substantial challenges facing rice breeding. We suggest
here that historical benchmarks be used in to evaluate promising rice
cultivars in addition to side-by-side comparisons to maintain
perspective of the absolute yield gains being made.

@@

CIs = posterior_interval(RES_full)[1:3,] * scl
CIs[1,] = round((CIs[1,] + center)/1000, 1)
CIs = round(CIs, 1)
med_eff = fixef(RES_full) * scl
med_eff[1] = round((med_eff[1] + center)/1000, 1)
med_eff = round(med_eff, 1)

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
                       "Release year (kg ha\\tsups{-1} year\\tsups{-1})",
                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

print(xtable(RES_coef,
             caption = "The estimated median fixed effects and 90\\% credible intervals for the improvement in yield of medium grain rice cultivars developed by the California Cooperative Rice Research Foundation, California, USA over the time period 1984 -- 2016.",
             label = "tbl:coef",
             align = "lcc"),
      caption.placement="top",
      booktabs=TRUE,
      sanitize.colnames.function = function(x){x},
      sanitize.rownames.function = function(x){x})
@
\newpage
